Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-05 article-models-tt.tex
added rule for families of families

\subsection{The basic ingredients of the type theory of models}
We first introduce the basic ingredients of our abstract theory of models.
\begin{align*}
& \inference
  {}
  {\jctx{\mctx}}
  \tag{the canonical model}\\
& \inference
  {\jfam{\mctx}{P}}
  {\jterm{\mctx}{P}{i}}
  \tag{initiality of $\mctx$}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jfam{{\Gamma}{A}}{\mfam{A}}}
  \tag{families}\\
& \inference
  {\jctx{\Gamma}}
  {\jfam{{\Gamma}{\mfam{\Gamma}}}{\mtm{\Gamma}}}
  \tag{terms}
\intertext{The following two rules essentially make a start with saying that every term is functorial
in the apropriate sense (there will be more rules contributing to this vies):}
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{\Gamma}{A}}{\mfam{A}}}{\subst{f}{\mfam{P}}}{\mfam{f}}}\\
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\subst{\mfam{f}}{{f}{\mtm{P}}}}{\mtm{f}}}
\intertext{The empty context of a model $A$ is simply a term of $A$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{\Gamma}{A}{\tfemp{A}}}
  \tag{empty context}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{{{\Gamma}{A}}{\mfam{A}}}{\subst{\tfemp{\mfam{A}}}{\mtm{A}}}{\tft{A}}}
\intertext{Extension is going to be a morphism from $\ctxext{A}{\mfam{A}}$ to
$A$ in context $\Gamma$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
  \tag{extension}
\intertext{The action of weakening on families
is a morphism from $\ctxwk{\mfam{A}}{\mfam{A}}$ to
$\mfam{\mfam{A}}$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{{{\Gamma}{A}}{\mfam{A}}}{\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}}
  \tag{weakening}
\intertext{The action of weakening on terms
is a morphism from $\ctxwk{\mfam{A}}{\mtm{A}}$ to $\mtm{\mfam{\mfam{A}}}$ over
$\tfwk{A}^0$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
    {{{\Gamma}{A}}{\mfam{A}}}
    {\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}
    {\ctxwk{\mfam{A}}{\mtm{A}}}
    {\mtm{\mfam{\mfam{A}}}}
    {\tfwk{A}^1}}
  \tag{weakening}
\intertext{The action of substitution by a term on families
is going to be a morphism from $\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}$ to 
$\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$
in context $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}}
  \tag{substitution}
\intertext{Likewise, the action of substitution by a term on terms of those
families is going to be a morphism from $\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}$
to $\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}$ over $\tfsubst{A}^0$ in context
$\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}}
  {\tfsubst{A}^1}}
  \tag{substitution}
\intertext{The identity functions are coded by}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm
   {{{\Gamma}{A}}{\mfam{A}}}
   {\subst{{\idfunc[\mfam{A}]}{\tfwk{A}^0}}{\mtm{\mfam{\mfam{A}}}}}
   {\tfid{A}}}
   \tag{identity functions}
\end{align*}
We will call the terms that we introduced here the model constructors.

\subsection{The compatibility rules}
We need to do several things in this section:
\begin{enumerate}
\item postulate that ordinary extension, weakening, substitution and identity
functions are compatible with the model constructors.
\item postulate that each of the model constructors is compatible with ordinary
extension, weakening, substitution and identity functions.
\item postulate that the model constructors are compatible with each other,
so that they come to model ordinary extension, weakening, substitution and
identity functions respectively.
\end{enumerate}
Note that the rules for compatibility with extension are going to explain which
model $\ctxext{A}{P}$ is by telling what the families, the terms, extension,
weakening, substitution and identity functions are. Likewise the rules for
compatibility with weakening and substitution do this for their respective
cases.

@@

@@

>>>>

\subsection{The basic ingredients of the type theory of models}
We first introduce the basic ingredients of our abstract theory of models.
\begin{align*}
& \inference
  {}
  {\jctx{\mctx}}
  \tag{the canonical model}\\
& \inference
  {\jfam{\mctx}{P}}
  {\jterm{\mctx}{P}{i}}
  \tag{initiality of $\mctx$}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jfam{{\Gamma}{A}}{\mfam{A}}}
  \tag{families}\\
& \inference
  {\jctx{\Gamma}}
  {\jfam{{\Gamma}{\mfam{\Gamma}}}{\mtm{\Gamma}}}
  \tag{terms}
\intertext{%
  The following two rules essentially make a start with saying that every term is functorial
  in the apropriate sense (there will be more rules contributing to this vies):}
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jhom
    {{\Gamma}{A}}
    {\mfam{A}}
    {\subst{f}{\mfam{P}}}
    {\mfam{f}}}\\
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jfhom
    {{\Gamma}{A}}
    {\mfam{A}}
    {\subst{f}{\mfam{P}}}
    {\mfam{f}}
    {\mtm{A}}
    {\subst{f}{\mtm{P}}}
    {\mtm{f}}}
\intertext{%
  When the situation requires clarity --  for instance when extensions
  are involved -- we will write $\mfam{f}^\Gamma$ and $\mtm{f}^\Gamma$ to make
  what we regard as the context explicit.}
\intertext{%
The empty context of a model $A$ is simply a term of $A$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{\Gamma}{A}{\tfemp{A}}}
  \tag{empty context}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{{{\Gamma}{A}}{\mfam{A}}}{\subst{\tfemp{\mfam{A}}}{\mtm{A}}}{\tft{A}}}
\intertext{%
  Extension is going to be a morphism from $\ctxext{A}{\mfam{A}}$ to
  $A$ in context $\Gamma$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
  \tag{extension}
\intertext{%
  The action of weakening on families
  is a morphism from $\ctxwk{\mfam{A}}{\mfam{A}}$ to
  $\mfam{\mfam{A}}$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{{{\Gamma}{A}}{\mfam{A}}}{\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}}
  \tag{weakening}
\intertext{%
  The action of weakening on terms
  is a morphism from $\ctxwk{\mfam{A}}{\mtm{A}}$ to $\mtm{\mfam{\mfam{A}}}$ over
  $\tfwk{A}^0$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
    {{{\Gamma}{A}}{\mfam{A}}}
    {\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}
    {\ctxwk{\mfam{A}}{\mtm{A}}}
    {\mtm{\mfam{\mfam{A}}}}
    {\tfwk{A}^1}}
  \tag{weakening}
\intertext{%
  The action of substitution by a term on families
  is going to be a morphism from $\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}$ to 
  $\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$
  in context $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}}
  \tag{substitution}
\intertext{%
  Likewise, the action of substitution by a term on terms of those
  families is going to be a morphism from $\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}$
  to $\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}$ over $\tfsubst{A}^0$ in context
  $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}}
  {\tfsubst{A}^1}}
  \tag{substitution}
\intertext{%
  The identity functions are coded by}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm
   {{{\Gamma}{A}}{\mfam{A}}}
   {\subst{{\idfunc[\mfam{A}]}{\tfwk{A}^0}}{\mtm{\mfam{\mfam{A}}}}}
   {\tfid{A}}}
   \tag{identity functions}
\end{align*}
We will call the terms that we introduced here the model constructors.

\subsection{The compatibility rules}
We need to do several things in this section:
\begin{enumerate}
\item postulate that ordinary extension, weakening, substitution and identity
functions are compatible with the model constructors.
\item postulate that each of the model constructors is compatible with ordinary
extension, weakening, substitution and identity functions. Actually, we want
that every term of this type theory is compatible with those. I.e.~every term
is a functor/morphism of models. Also the sections.
\item postulate that the model constructors are compatible with each other,
so that they come to model ordinary extension, weakening, substitution and
identity functions respectively.
\end{enumerate}
Note that the rules for compatibility with extension are going to explain which
model $\ctxext{A}{P}$ is by telling what the families, the terms, extension,
weakening, substitution and identity functions are. Likewise the rules for
compatibility with weakening and substitution do this for their respective
cases.

@@

\subsubsection{Families over families are families over extensions}
\begin{equation*}
\inference
  {\jfam{\Gamma}{A}}
  {\jfameq
    {{{\Gamma}{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\jcomp{{A}{\mfam{A}}}{\tfext{A}}{\mfam{A}}}}
\end{equation*}
\emph{(Note: this rule might be a consequence of an explanation of what $\mfam{{A}{P}}$
is in general, but at the moment I don't see how to do this)}

\subsubsection{Extension acts as the identity on families of families}
The family of which $\tfext{A}$ is a term, is unfolded as
\begin{equation*}
\unfold{\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
\end{equation*}
Therefore, there is a term
\begin{equation*}
\jhom
  {{\Gamma}{A}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
By the rule asserting that families of families are families of extensions, we
have $\mfam{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mfam{A}}$. By the rule that
families over weakenings are weakenings of families, we have the judgmental
equalities
\begin{equation*}
\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq \unfold{\jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}}
\jdeq \jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}
\end{equation*}
Therefore, we can compare the term $\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}$
to the identity function $\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$. 
The compatibility
rule for the action of extension on families of families asserts that
\begin{equation*}
\inference
  {\jfam{\Gamma}{A}}
  {\jhomeq
    {{\Gamma}{A}}
    {\jcomp{}{\tfext{A}}{\mfam{A}}}
    {\jcomp{}{\tfext{A}}{\mfam{A}}}
    {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
    {\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]}}
\end{equation*}
\emph{(Note: this rule might be a consequence of an explanation of what $\mfam{{A}{P}}$
is in general, but at the moment I don't see how to do this)}

@@

\subsubsection{Extension acts as the identity functor on families}
\begin{equation*}
df
\end{equation*}

\subsubsection{Extension acts as the identity functor on terms}
